---
title: "AgLand"
author: "Mandy Liesch"
date: "3/17/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Purpose

The purpose of this vignette is looking at the land use changes around the globe by different land use types using  the R Terra Package, masking, zonal statisitics, and visualizations with bivariate cholopleth maps to visualize the cropland gains and losses by country/state. 

A Note on file size:

`terraOptions(tempDir="F:/temp")`

## R Packages
```{r}

library(biscale)
library(cowplot)
library(exactextractr)
library(gdalUtils)
library(geodata)
library(ggplot2)
library(ggridges)
library(rgdal)
library(sf)
library(terra)
library(tidyverse)

terraOptions(tempDir="F:/temp")
```


## Raw Data and Preparations 

### Countries

```{r}

w <- world(path=tempdir())

plot(w)

w

```



### Soil Grids: Total Soil Carbon Stocks 0-30 cm

[SoilGrids](soilgrids.org) is a system for global digital soil mapping that makes use of global soil profile information and covariate data to model the spatial distribution of soil properties across the globe. SoilGrids is a collections of soil property maps for the world produced using machine learning at 250 m resolution. The [Soil Grids 2.0 FAQs](https://www.isric.org/explore/soilgrids/faq-soilgrids#How_can_I_access_SoilGrids) details the file structure and downloads for both the soil grids data for soil carbon, as well as the uncertainty.

This code process takes you through how to grab the WebDAV protocol modelled off of the [SoilGrids Notebook](https://git.wur.nl/isric/soilgrids/soilgrids.notebooks/-/blob/master/markdown/webdav_from_R.md), using the GDAL functionality in R.



```{r, eval=FALSE}
library(rgdal)
library(gdalUtils)

link="/vsicurl/https://files.isric.org/soilgrids/latest/data/ocs/ocs_0-30cm_Q0.5.vrt"

gdal_translate(link,
    "soilC_Rast.tif",
    co=c("TILED=YES","COMPRESS=DEFLATE","BIGTIFF=YES"),
    verbose=TRUE)



```


```{r, eval=FALSE}

scStock<-rast("soilC_Rast.tif")

plot(scStock)

project(scStock, w, filename="Reproj_SoilC.tif")

```

```{r}

scStock<-rast("Reproj_SoilC.tif")

plot(scStock)
plot(w, bg="transparent", add=TRUE)

```


### Cropland Extent Changes

The dataset represents a globally consistent cropland extent time-series at 30-m spatial resolution. Cropland defined as land used for annual and perennial herbaceous crops for human consumption, forage (including hay), and biofuel. It is on the [Global Land Analysis and Discovery](https://glad.umd.edu/dataset/croplands), on a project through the University of Maryland. 

The chosen rasters downloaded are the Cropland Gain (Global_cropland_3km_netgain.tif).

The net cropland extent change from 2003 to 2019 is shown as the Pixel value: 0-100, percent of cropland dynamic (loss or gain) per pixel.

```{r, eval=FALSE}

cropGain<-rast("Global_cropland_3km_netgain.tif")


project(cropGain, scStock, filename="Reproj_cropGain.tif")

```


```{r}
gains<-rast("Reproj_cropGain.tif")

plot(gains)
plot(w, bg="transparent", add=TRUE)

```


## Preliminary Analysis

```{r}

shapeSF <- sf::st_as_sf(w)

regions<-read_csv("regions.csv")

shapes<-merge(shapeSF, regions, by="GID_0")

shapes$TotalCount<-exact_extract(scStock, shapes, 'count')
shapes$totC_Mean<-exact_extract(scStock, shapes, 'mean')
shapes$cropGains_Mean<-exact_extract(gains, shapes, 'mean')


```

### Tables by Region

```{r}

summary<-na.omit(shapes) 
summary$geometry<-NULL


sumStats <- summary %>%
  group_by(Region) %>%
  summarise_at(vars(totC_Mean, cropGains_Mean), list(name = mean))

ggplot(summary, aes(x = cropGains_Mean, y = Region)) +
  geom_density_ridges(scale = 4) + 
  scale_y_discrete(expand = c(0, 0)) +     # will generally have to set the `expand` option
  scale_x_continuous(expand = c(0, 0)) +   # for both axes to remove unneeded padding
  coord_cartesian(clip = "off") + # to avoid clipping of the very top of the top ridgeline
  theme_ridges()



