---
title: "Global Cropland Expansion and Soils at Risk"
subtitle: "A Terra Tutorial"
author: "Mandy Liesch"

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = TRUE)
```

```{r, include=FALSE, eval=FALSE}
rmarkdown::render("globalAgLand.Rmd", 
                  output_format = "github_document", 
                  output_options = list(toc=TRUE, toc_depth = 4, html_preview=FALSE), 
                  output_file = "README.md")

```

# Purpose

The purpose of this vignette is looking at the land use changes around the globe by different land use types using  the R Terra Package, masking, zonal statisitics, and visualizations with bivariate cholopleth maps to visualize the cropland gains and losses by country/state. 

A Note on file size:

`terraOptions(tempDir="F:/temp")`

# R Packages
```{r}

library(biscale)
library(cowplot)
library(exactextractr)
library(gdalUtils)
library(geodata)
library(ggplot2)
library(ggridges)
library(rgdal)
library(rmarkdown)
library(sf)
library(terra)
library(tidyverse)

terraOptions(tempDir="F:/temp")
```


# Raw Data and Preparations 

## Countries

```{r}

w <- world(path=tempdir())

plot(w)

w

```



## Soil Grids: Total Soil Carbon Stocks 0-30 cm

[SoilGrids](soilgrids.org) is a system for global digital soil mapping that makes use of global soil profile information and covariate data to model the spatial distribution of soil properties across the globe. SoilGrids is a collections of soil property maps for the world produced using machine learning at 250 m resolution. The [Soil Grids 2.0 FAQs](https://www.isric.org/explore/soilgrids/faq-soilgrids#How_can_I_access_SoilGrids) details the file structure and downloads for both the soil grids data for soil carbon, as well as the uncertainty.

This code process takes you through how to grab the WebDAV protocol modelled off of the [SoilGrids Notebook](https://git.wur.nl/isric/soilgrids/soilgrids.notebooks/-/blob/master/markdown/webdav_from_R.md), using the GDAL functionality in R.



```{r, eval=FALSE}
library(rgdal)
library(gdalUtils)

link="/vsicurl/https://files.isric.org/soilgrids/latest/data/ocs/ocs_0-30cm_Q0.5.vrt"

gdal_translate(link,
    "soilC_Rast.tif",
    co=c("TILED=YES","COMPRESS=DEFLATE","BIGTIFF=YES"),
    verbose=TRUE)



```


```{r, eval=FALSE}

scStock<-rast("soilC_Rast.tif")

plot(scStock)

project(scStock, w, filename="Reproj_SoilC.tif")

```

```{r}

scStock<-rast("Reproj_SoilC.tif")

plot(scStock)
plot(w, bg="transparent", add=TRUE)

```


## Cropland Extent Changes

The dataset represents a globally consistent cropland extent time-series at 30-m spatial resolution. Cropland defined as land used for annual and perennial herbaceous crops for human consumption, forage (including hay), and biofuel. It is on the [Global Land Analysis and Discovery](https://glad.umd.edu/dataset/croplands), on a project through the University of Maryland. 

The chosen rasters downloaded are the Cropland Gain (Global_cropland_3km_netgain.tif).

The net cropland extent change from 2003 to 2019 is shown as the Pixel value: 0-100, percent of cropland dynamic (loss or gain) per pixel.

```{r, eval=FALSE}

cropGain<-rast("Global_cropland_3km_netgain.tif")


project(cropGain, scStock, filename="Reproj_cropGain.tif")

```


```{r}
gains<-rast("Reproj_cropGain.tif")

plot(gains)
plot(w, bg="transparent", add=TRUE)

```


# Preliminary Analysis

```{r}

shapeSF <- sf::st_as_sf(w)

regions<-read_csv("regions.csv")

shapes<-merge(shapeSF, regions, by="GID_0")

shapes$TotalCount<-exact_extract(scStock, shapes, 'count')
shapes$totC_Mean<-exact_extract(scStock, shapes, 'mean')
shapes$cropGains_Mean<-exact_extract(gains, shapes, 'mean')


```

## Tables by Region

```{r}

summary<-na.omit(shapes) 
summary$geometry<-NULL


sumStats <- summary %>%
  group_by(Region) %>%
  summarise_at(vars(totC_Mean, cropGains_Mean), list(name = mean))

```

```{r}
sumStats
```

## Ridgeplots of Regional Data

```{r}

cropGainPlot<-ggplot(summary, aes(x = cropGains_Mean, y = Region)) +
  geom_density_ridges(scale = 4) + 
  scale_y_discrete(expand = c(0, 0)) +     # will generally have to set the `expand` option
  scale_x_continuous(expand = c(0, 0)) +   # for both axes to remove unneeded padding
  coord_cartesian(clip = "off") + # to avoid clipping of the very top of the top ridgeline
  theme_ridges()


cropGainPlot

SOCPlot<-ggplot(summary, aes(x = totC_Mean, y = Region)) +
  geom_density_ridges(scale = 4) + 
  scale_y_discrete(expand = c(0, 0)) +     # will generally have to set the `expand` option
  scale_x_continuous(expand = c(0, 0)) +   # for both axes to remove unneeded padding
  coord_cartesian(clip = "off") + # to avoid clipping of the very top of the top ridgeline
  theme_ridges()

SOCPlot

```

## Bivariate Chloropleth Maps By Region

```{r}
shapes<-na.omit(shapes) 

shapes <- bi_class(shapes, x = cropGains_Mean, y = totC_Mean, style = "jenks", dim = 3)

map <- ggplot() +
  geom_sf(data = shapes, mapping = aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
  bi_scale_fill(pal = "DkBlue", dim = 3) +
  labs(
    title = "Cropland Gains and 30 cm Carbon Stocks",
  ) +
  bi_theme()

legend <- bi_legend(pal = "DkBlue",
                    dim = 3,
                    xlab = "Cropland Gain",
                    ylab = "Carbon Stocks",
                    size = 12)

finalPlot <- ggdraw() +
  draw_plot(map, 0, 0, 1, 1) +
  draw_plot(legend, 0, 0.2, 0.2, 0.2)


finalPlot

```

## Dyanamic Country Results

```{r}

newShape<-shapes
newShape$geometry<-NULL

highRisk<-newShape %>%
  filter(bi_class == '3-3'|bi_class == '3-2') %>%
  select(NAME_0, bi_class, Continent, totC_Mean, cropGains_Mean)

highRisk

```

# Detailed Analysis

```{r, eval=FALSE}

target<-gains>=10


targMask<-classify(
#define the raster you are clipping
target,
#define the reclassification values (id is 0= not in the 10% or more)
cbind(id=c(0, 1),
#these are the raster values, with an id of 0 returning as an NA, otherwise it is 1.
v=c(NA, 1)),
filename='targMask.tif')


```

```{r}
targetMask<-rast("targMask.tif")

```

```{r, eval=FALSE}
highCrop<-mask(
#the first raster is the full raster we are cutting (the soil carbon stocks)
scStock,
#the second raster is the mask we created in the previous step
targetMask,
#If TRUE, areas on mask that are _not_ the maskvalue are masked, since we classified them as NA, this can be set to false
inverse=FALSE,
#The value that cells of x should become if they are not covered by mask, set to NA here.
updatevalue=NA,
#new creation of a .tif file, with the cut values of soil carbon remaining.
filename="highCrop_SOC_Stocks.tif")

```

```{r}
highCropCh<-rast("highCrop_SOC_Stocks.tif")

plot(highCropCh)
plot(w, bg="transparent", add=TRUE)

```


```{r}
shapes$hCCount<-exact_extract(highCropCh, shapes, 'count')
shapes$hcMean<-exact_extract(highCropCh, shapes, 'mean')

shapes$perChange<-(shapes$hCCount/shapes$TotalCount)*100

ggplot(data = shapes) +
    geom_sf(aes(fill = perChange)) +
    scale_fill_viridis_c(option = "plasma")
```

```{r}
shapes2<-na.omit(shapes) 

shapes2 <- bi_class(shapes2, x = perChange, y = hcMean, style = "jenks", dim = 3)

map <- ggplot() +
  geom_sf(data = shapes2, mapping = aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
  bi_scale_fill(pal = "DkBlue", dim = 3) +
  labs(
    title = "High Risk Countries and Carbon",
  ) +
  bi_theme()

legend <- bi_legend(pal = "DkBlue",
                    dim = 3,
                    xlab = "High Change Areas",
                    ylab = "Carbon Stocks",
                    size = 12)

finalPlot2 <- ggdraw() +
  draw_plot(map, 0, 0, 1, 1) +
  draw_plot(legend, 0, 0.2, 0.2, 0.2)


finalPlot2

```

```{r}
newShape1<-shapes2
newShape1$geometry<-NULL

highRisk<-newShape1 %>%
  filter(bi_class == '3-3'|bi_class == '3-2'|bi_class == '2-3'|bi_class == '3-3') %>%
  select(NAME_0, bi_class, Continent, totC_Mean, perChange)

highRisk

```
